Scalable Clustering of Genotype Information using MapReduce

### 2014-05-19 doc/publications/incob2014/paper.tex
Updated

% SCALING UP FEATURES IN MAHOUT
\subsection*{Mahout scales to 15 Million genomic variants}
In this section we demonstrate Mahout's scaling ability by increasing the numbers of features, here variants per sample. 
Including more regions for the similarity calculation will increase the cluster performance, however also increases the memory consumption. 
R therefore was not able to perform the clustering on even the medium sized dataset (\SixReal , 15 Million variants), we hence only report resource consumption and accuracy for Mahout.
On this dataset Mahout's pre-processing takes 325 minutes, with 13GB of physical and 37GB of virtual memory, and the clustering takes 80 minutes, with XXGB of physical and XXGB of virtual memory. 
The clustering accuracy has increased by 35\% to 0.812 with 10 clusters compared to the performance on \OneReal\ for the 4-group-based population task (see Table~\ref{datasetsAcc}).

@@

% SCALING UP SAMPLES IN MAHOUT
\subsection*{Mahout scales effortless to an order of magnitude more samples}
In this section we investigate Mahout's ability to scale with more samples.
The resource consumption remains virtually identical between \SixReal\ and the order of magnitude larger dataset \SixArtiExact, which takes XX minutes, with XXGB of physical and XXGB of virtual memory, and the clustering takes XX minutes, with XXGB of physical and XXGB of virtual memory. 
The clustering accuracy is YY with XX clusters.

@@

\section*{Conclusion}
In this paper we have applied Mahout's k-means clustering to the task of genotype based population and family association. 
We provide an interface from VCF format to Mahout and a toolset to interpret Mahout's clustering output (see Figure~\ref{fig:sign})
We find that on small datasets the actual clustering stage is two orders of magnitudes slower in Mahout compared to R (10 minutes vs 7 seconds), likely due to the overhead of setting up and co-ordinating the communication between the Hadoop nodes. 
However, increasing either the number of samples or the number of features per sample quickly exceeds the ability of R.
Mahout on the other hand scales effortlessly to five times more samples (5,460) and copes well with 15 Million features per samples.
As memory consumption directly scales with the number of features per sample, we observe that Mahout's standard set-up of Hadoop does not scale to a whole exome dataset (38 Million features per sample).
However, this might be resolved once Mahout moves from MapReduce to a domain-specific language for linear algebraic operations on top of Apache Spark \url{http://spark.apache.org/}.

>>>>

% SCALING UP FEATURES IN MAHOUT
\subsection*{Mahout scales to 15 Million genomic variants}
In this section we demonstrate Mahout's scaling ability by increasing the numbers of features, here variants per sample. 
Including more regions for the similarity calculation will increase the cluster performance, however also increases the memory consumption. 
R therefore was not able to perform the clustering on even the medium sized dataset (\SixReal , 15 Million variants), we hence only report resource consumption and accuracy for Mahout.
On this dataset Mahout's pre-processing takes 325 minutes, with 13GB of physical and 37GB of virtual memory, and the clustering takes 80 minutes.
The clustering accuracy has increased by 35\% to 0.812 with 10 clusters compared to the performance on \OneReal\ for the 4-group-based population task (see Table~\ref{datasetsAcc}).

@@

% SCALING UP SAMPLES IN MAHOUT
\subsection*{Mahout scales effortless to an order of magnitude more samples}
In this section we investigate Mahout's ability to scale with more samples.
The resource consumption remains virtually identical between \SixReal\ and the order of magnitude larger dataset \SixArtiExact, which takes 33 minutes, with 6GB of physical and 39GB of virtual memory, and the clustering takes 23 minutes 
The clustering accuracy is 0.812 with 10 clusters.

@@

\section*{Conclusion}
In this paper we have applied Mahout's k-means clustering to the task of genotype based population and family association. 
We provide an interface from VCF format to Mahout and a toolset to interpret Mahout's clustering output (see Figure 1)
We find that on small datasets the actual clustering stage is two orders of magnitudes slower in Mahout compared to R (10 minutes vs 7 seconds), likely due to the overhead of setting up and co-ordinating the communication between the Hadoop nodes. 
However, increasing either the number of samples or the number of features per sample quickly exceeds the ability of R.
Mahout on the other hand scales effortlessly to five times more samples (5,460) and copes well with 15 Million features per samples.
As memory consumption directly scales with the number of features per sample, we observe that Mahout's standard set-up of Hadoop does not scale to a whole exome dataset (38 Million features per sample).
However, this might be resolved once Mahout moves from MapReduce to a domain-specific language for linear algebraic operations on top of Apache Spark \url{http://spark.apache.org/}.
